---
title: "Probabilistic Spatial Partition Model"
author: "Carl Müller-Crepon"
output:
   md_document:
      variant: gfm

---


# pspm: Probabilistic Spatial Partition Model

## Introduction

This packages contains a series of functions and classes to set up and fit a Probabilistic Spatial Partition Model and sample partitionings following the the methodology introduced by Müller-Crepon, Schvitz and Cederman (2023). In short, the model is based on a simplification of continuous geographic space as a (planar) graph. The observed partitioning is encoded on the graphs' vertices such that each vertex is a member of one and only one partition. The graphs partitioning is modeled as the result of _attractive_ and _repulsive_ force active on its edges which affect the probability that the vertices they connect belong to the same or to different partitions. The model allows to estimate the effects of edge-elvel attributes on the repulsion/attraction between vertices, thus estimating their impact on the overall partitioning of the graph.  

Please refer to the original publication for all technical details beyond the summary provided below. It can be found [here]() and as ungated version [here](http://www.carlmueller-crepon.org/publication/state_shape/). 

When using the pspm package, please cite:

Müller-Crepon, Carl, Guy Schvitz, Lars-Erik Cederman (2023).
Shaping States into Nations: The Effects of Ethnic Geography on State Borders.
_American Journal of Political Science_, conditionally accepted for publication.  


## The model
We model the distribution over all possible partitionings $P$ of lattice $G$ as a Boltzmann distribution:

```math
Pr(P = p_{i}) =  {e^{-\epsilon_{i}}\over\displaystyle\sum_{i = 1}^{|\mathbb{P}|}e^{-\epsilon_{i}} }
```

here, a partitioning $i$'s chance of realization decreases with its energy $\epsilon_i$. The energy of a partitioning is defined as the sum of energies on the graph's edges that run between nodes $j$ and $k$ where $j$ and $k$ are located in the same partition ($s_{j,k} = 1$). Edges energies are not realized where $j$ and $k$ do not belong to the same partition ($s_{j,k} = 0$):

```math
\epsilon_{i} = \displaystyle\sum_{j,k \in L}\, \epsilon_{j,k}\,s_{j,k}
```

Potential and realized edge-level energies are in turn affected by a constant ($\beta_0$) and a set of edge-level attributes $\textbf{x}_{j,k}$ that are weighted by a parameter vector $\beta$

```math
\epsilon_{j,k} = \beta_0 + \beta\, \textbf{x}_{j,k}
```

The empirical goal of the PSPM is it to estimate parameters $\beta_0$ and $\beta$ from observed data. To that intent, the PSPM uses a maximum composite likelihood approach. Uncertainty estimates can be derived via a parametric bootstrap that samples partitionings with a given set of $\beta_0$ and $\beta$ parameters and then fits the model on the sampled partitionings. 



## Installation

You can directly download and install the pspm package from GitHub. Before doing so, please make sure that you have [Python3](https://www.python.org/downloads/) installed. Upon installation, the package automatically installs necessary python dependencies via the [reticulate](https://cran.r-project.org/web/packages/reticulate/index.html) R-package. 

```{r, eval = F}
library(devtools)
install_github(repo = "carl-mc/pspm")
```


## Getting started

```{r, eval = T, warning=FALSE}
# Packages
library(pspm)
library(igraph)

# Set seeds in R and python
pspm_set_seed(1)

```


## Setting up and handling a toy lattice

```{r, eval = T}
# Make mock PSPM Object with a sampled partitioning
sl <- generate_grid_data(N_sqrd = 10, ## 10 x 10 lattice
                        beta0 = -2, ## Negative constant = baseline attraction between nodes
                        beta = c(2,1), ## Include two repulsive edge-level predictors
                        dep_structure = "von_neumann", ## Each node connects to 4 neighbors
                        burnin = 10 ## Sample with a 10 burn-in periods
                        )
sl$plot_partitioning(edge_predictor = 1, 
                     edge.width = 5, vertex.size = 10,
                     main = "A Toy Lattice")
```


## PSPM to igraph conversion

```{r, eval = T}
# Transform PSPM to igraph
graph <- PSPM2igraph(sl)
edge_attr_names(graph)
vertex_attr_names(graph)

# Transform igraph back to PSPM
sl.from.g <- igraph2PSPM(g = graph, outcome_name = "Y",
                         edge_pred_names = c("x1", "x2"))

```


## Fitting a PSPM Model

```{r, eval = T}
## The easy way

### Estimate
m.simple <- fit_pspm_model(formula = Y ~ x1 + x2, 
                           g_ls = list(graph),
                           return_pspm = TRUE)

# ### Bootstrap CIs
# bs.simple <- bootstrap_pspm(m.simple, 
#                             n_boot_iter = 10, ## Should be > 100
#                             burnin = 10, ## Could be higher, depending on complexite of graph and model
#                             cl = 10L, ## Number of CPUs for parallelization
#                             return_sims = TRUE, ## Return full distribution of estimates
#                             ci_level = .95)
# 
# ### Summary
# summary(m.simple)
# print(bs.simple$ci_mat)
# plot(density(bs.simple$beta_boot[,"x1"]),
#      main = "Distribution of bootstrapped estimates of x1")

```




```{r, eval = T}
## The complicated way (inside the wrapper)

### Initiate PSPMLearn Object
learn_obj <- PSPMLearn$new(list(sl.from.g))

# ### Fit
# m.compl <- learn_obj$fit_composite_log_likelihood(beta_init = c(0,0,0))
# 
# ### Bootstrap
# bs.compl <- learn_obj$par_bootstrap_composite_log_likelihood(n_boot_iter = 10, burnin = 10, 
#                                                  cl = 10L, return_sims = FALSE, ci_level = .95)
# 
# ### Summary
# summary(m.compl)
# print(bs.compl)
```


## Producing nice tables
```{r, eval = T}
# Integration with texreg
pspm2table(list(m.simple),
           # bootci = list(bs.simple$ci_mat), boottype = "percentile",
           type = "text", add.stats = c("Edges" = "N_edges", "Vertices" = "N"))


```